---
title: "Overdose Death Project using SCC Death Data"
author: "Dawson Coblin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Meeting and Project Notes  

### (12/1/24): Current Objectives  

Following our meeting on December 1, 2024, I located the coroners dashboard and 
downloaded the overdose death data for Santa Clara County. I emailed Mile P., to 
confirm the adequateness for use with r-programming visualization purposes.

The link used for the data set is:
https://data.sccgov.org/Health/Medical-Examiner-Coroner-Total-Drug-Deaths-dataset/9v2c-rxbj/data

Deliverable: (deliverable due: 12/16/24)
I will use this data set to review and form age groups (current), to investigate 
and plot points and formulate theories that will use various statistical models.

PPDAC Model (TBD)
Problem:
Plan:
Data:
Analysis:
Conclusion:

### (11/2/24): Revised to 12/1/24 objectives

Hi Wayne!  

Thank you again for taking the time to review actual data with me, while helping
to foster those epidemiological thought processes with that data!
Following our Philz meeting, I took some time to review the various data sources 
used to compile and generate the SCC's dashboard for overdose death data. I found 
that SCC's death data is sourced from California's Vital Records Business 
Information System (VRBIS). I hit a roadblock in finding a comprehensive data set 
for overdose death data. After venturing through VRBIS, I, unfortunately, could 
not access the death data used to generate the SCC dashboard.  
However, looking through VRBIS, I stumbled across the California Vital Data 
(Cal-ViDa) Query Tool, which allowed me to create limited characteristic data 
sets for overdose data that is specific to SCC. I have those files in the 
repository.  

Given the restricted use of the data sets and the flexibility with the 
objectives at hand, I took the opportunity to brainstorm theoretical project ideas. 
I hope you don't mind!  

At the very least, they may be interesting discussion points - Please let me 
know your thoughts!  

### 1. Geospatial Analysis of Overdose Death Hot-spots  

#### Objective:  
Identify geographic hot-spots (zip code level?) within the county for
overdose deaths.  

#### Focus:  
Determine whether certain areas (e.g., urban vs. rural, economically
disadvantaged neighborhoods) have higher rates of overdose deaths.  

#### Methods:  
Use GIS mapping to visualize overdose death rates across different
areas in the county, Use those findings to use for other projects or possibly
provide guidance information to the SCC's overdose coalition group to encourage
program improvements.  

### 2. Comparing Demographic Risk Factors  

#### Objective:  
Investigate demographic characteristics (age, gender, race/ethnicity)
linked to higher overdose death rates.  

#### Focus:  
Determine which demographic groups are most affected by overdose deaths,
identifying potential vulnerabilities.  

#### Methods:  
Use stratified analysis to calculate rates within demographic
subgroups, highlighting significant differences and exploring social
determinants of health.  

### 3. Association with Access to SUD Treatments  

#### Objective:  
Examine how access to SCC opioid and other drug assistance programs for
overdose help and community-provided services at county facilities that offer
harm reduction programs correlates with overdose death rates.  

#### Focus:  
Investigate if regions with limited access to substance abuse services
have higher overdose death rates.

#### Methods:  
Map overdose deaths alongside treatment facility locations and perform
correlation analysis.  

\newpage


\newpage  

## Included State DataSets

```{r message=FALSE, warning=FALSE}
library(readr)
library(tidyr)
library(dplyr)
library(tinytex)
library(ggplot2)
```

```{r Datasets, message=FALSE, warning=FALSE, paged.print=FALSE}
#Post 12/1/24 meeting: Data set exported from SCC dashboard #(https://data.sccgov.org/Health/Medical-Examiner-Coroner-Total-Drug-Deaths-dataset/9v2c-rxbj/data).

Medical_Examiner_Coroner_Total_Drug_Deaths_dataset <- read_csv("~/_Personal/Epi work/SCC OD Death Data Project_Wayne/NEW OD Death Project_Harddrive/OD-Death-Data-SCC_-main/Medical_Examiner-Coroner__Total_Drug_Deaths_dataset.csv")

View(Medical_Examiner_Coroner_Total_Drug_Deaths_dataset)
```

```{r}
#Creating age groups using group by and aggregate:

Age_Brackets_name_1 <- Medical_Examiner_Coroner_Total_Drug_Deaths_dataset %>%
  mutate(Age_Brackets = cut(Age, breaks = c(0, 20, 30, 40, 50, 60, 70, 80, 90),
                            labels = c("0-20", "21-30", "31-40", "41-50",
                                       "51-60", "61-70", "71-80", "81-90")))



Age_Brackets_Table<- 
  Age_Brackets_name_1 %>% group_by(Age_Brackets) %>%
  summarise(count = n(), .groups = "drop")

View(Age_Brackets_Table)

ggplot(data = Age_Brackets_Table, mapping = aes(x= Age_Brackets, 
                                                y= count))
```
